Fasano--Franceschini Test: an Implementation of a 2-Dimensional Kolmogorov--Smirnov test in \proglang{R}

### 2021-06-17 journalMS/article.tex
fixed spelling and grammar mistakes

%% - \Abstract{} almost as usual
\Abstract{
The univariate Kolmogorov-Smirnov (KS) test is a non--parametric
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample
case,  whether the two
samples come from the same underlying distribution).
The versatility of the KS test has
made it a cornerstone of statistical analysis and is commonly used
across the scientific disciplines. However, the test proposed by
Kolmogorov and Smirnov does not naturally extend to multidimensional
distributions.  Here, we present the \pkg{fasano.franceschini.test}
package, an  \proglang{R} implementation of the 2-D KS two--sample
test as defined by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides three improvements
over the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; (ii) the package implements a procedure for handling ties in
the data; and (iii) the package implements a parallelized
bootstrapping procedure for improved significance testing. Ultimately,
the \pkg{fasano.franceschini.test} package presents a robust
statistical test for analysing random samples defined in 2-dimensions.
}

@@

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered on each observed $(x, y)$ pair to compute the
cumulative distribution functions. That is, rather than looking over all $n^2$ points
${(X_i, Y_j): i,j \in [1,n]}$, Fasano and Franceschini only use the observed $n$ points
${(X_i, Y_i): i \in [1,n]}$.
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

To assess the computational efficiency, we benchmarked the package as follows.
Using the \pkg{rbenchmark} package to evaluate runtime, the Fasano--Franceschini
test and Peacock test were run under four different samples sizes
($n=10, 100, 1000, 5000$), with 10 replicates for each run. The
Fasano--Franceschini test bootstrap precedure was further evaluated
under four different bootstrap iterations (no bootstrap, 10, 100, 1000),
again using 10 replicates for each run. Reported results represent the
average run time of the 10 replicate benchmarks. All benchmark tests were run on a 2018 macBook
Pro Mac (macOS Catalina) with a 2.7-GHz Quad-Core Intel Core i7 processor and 16 GB of 2133 MHz LPDDR3 memory.

The main distinction between the Peacock and Fasano--Franceschini
tests is in computational efficiency, with  Fasano--Franceschini
scaling as $O(n^2)$ relative to Peacock's complexity of  $O(n^3)$
\citep{Lopes2007}.  Our benchmarks also show this advantage, as shown
in \textbf{Figure~\ref{fig:bmark}A}.  While the implentation of the
bootstrapping procedure increases runtime in comparison to the
approximate fit from  \citet{numericalRecipes}, parallelization of the
Fasano--Franceschini test shows a four-fold reduction in run time when
parallelized across 8 cores (\textbf{Figure~\ref{fig:bmark}B}).
% (See the \pkg{Peacock.test} package for \proglang{R} implementation of the Peacock test.)

>>>>

%% - \Abstract{} almost as usual
\Abstract{
The univariate Kolmogorov-Smirnov (KS) test is a non--parametric
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample
case,  whether the two
samples come from the same underlying distribution).
The versatility of the KS test has
made it a cornerstone of statistical analysis and is commonly used
across the scientific disciplines. However, the test proposed by
Kolmogorov and Smirnov does not naturally extend to multidimensional
distributions.  Here, we present the \pkg{fasano.franceschini.test}
package, an  \proglang{R} implementation of the 2-D KS two--sample
test as defined by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides three improvements
over the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; (ii) the package implements a procedure for handling ties in
the data; and (iii) the package implements a parallelized
bootstrapping procedure for improved significance testing. Ultimately,
the \pkg{fasano.franceschini.test} package presents a robust
statistical test for analyzing random samples defined in 2-dimensions.
}

@@

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered on each observed $(x, y)$ pair to compute the
cumulative distribution functions. That is, rather than looking over all $n^2$ points
${(X_i, Y_j): i,j \in [1,n]}$, Fasano and Franceschini only use the observed $n$ points
${(X_i, Y_i): i \in [1,n]}$.
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The procedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

To assess the computational efficiency, we benchmarked the package as follows.
Using the \pkg{rbenchmark} package to evaluate runtime, the Fasano--Franceschini
test and Peacock test were run under four different samples sizes
($n=10, 100, 1000, 5000$), with 10 replicates for each run. The
Fasano--Franceschini test bootstrap procedure was further evaluated
under four different bootstrap iterations (no bootstrap, 10, 100, 1000),
again using 10 replicates for each run. Reported results represent the
average run time of the 10 replicate benchmarks. All benchmark tests were run on a 2018 macBook
Pro Mac (macOS Catalina) with a 2.7-GHz Quad-Core Intel Core i7 processor and 16 GB of 2133 MHz LPDDR3 memory.

The main distinction between the Peacock and Fasano--Franceschini
tests is in computational efficiency, with  Fasano--Franceschini
scaling as $O(n^2)$ relative to Peacock's complexity of  $O(n^3)$
\citep{Lopes2007}.  Our benchmarks also show this advantage, as shown
in \textbf{Figure~\ref{fig:bmark}A}.  While the implementation of the
bootstrapping procedure increases runtime in comparison to the
approximate fit from  \citet{numericalRecipes}, parallelization of the
Fasano--Franceschini test shows a four-fold reduction in run time when
parallelized across 8 cores (\textbf{Figure~\ref{fig:bmark}B}).
% (See the \pkg{Peacock.test} package for \proglang{R} implementation of the Peacock test.)
